---
layout: post
comments: true
title: Scale Only the Continuous Variables in an R Data Frame.
---

# Scale Only the Continuous Variables in an R Data Frame.

When performing a logistic regression, often your dataset consists of a mixture of continuous and binary variables. In order to avoid numerically-unstable estimation, it is desirable to scale the continous variables in your dataset while leaving the binary variables alone. Consider the following data frame:

```{r}
df = data.frame(x = c(rep(0, 5), rep(1, 5)), 
                y = 11:20, 
                z = 21:30)

df
```

To obtain a logical vector indicating which columns are binary, we can use `apply()`:

```{r}
binary = apply(df, 2, function(x) {all(x %in% 0:1)})

binary
```

`apply()` found all columns whose elements consist only of 0 and 1. Now we can subset the data using this logical vector. We want the continuous variables, though, so we use `!`:

```{r}
df[!binary]
```

Finally, we can write a function that scales only the continuous variables in a data frame.

```{r}
scaleContinuous = function(data) {
  binary = apply(data, 2, function(x) {all(x %in% 0:1)}) 
  data[!binary] = scale(data[!binary])
  return(data)
}

scaleContinuous(df)
```