---
title: "Visualize Your Graph with RNeo4j and visNetwork."
output: html_document
layout: post
---

```{r, echo=FALSE}
library(RNeo4j)
graph = startGraph("http://localhost:7474/db/data/")
clear(graph, input=F)
```

Someone recently [posted an issue](https://github.com/nicolewhite/RNeo4j/issues/28) on RNeo4j where they needed help visualizing their graph in [visNetwork](https://github.com/dataknowledge/visNetwork), which has turned out to be a pretty fun R package. I decided to turn my answer into a blog post.

Both RNeo4j and visNetwork need to be installed with `devtools`:

```{r, eval=FALSE}
devtools::install_github("nicolewhite/RNeo4j")
devtools::install_github("dataknowledge/visNetwork")
```

```{r, message=FALSE}
library(RNeo4j)
library(visNetwork)
```

Then, with [Neo4j](http://neo4j.com) installed and running, we can connect to our graph and create some data:

```{r, results='hide'}
graph = startGraph("http://localhost:7474/db/data/")

nicole = createNode(graph, "Person", name="Nicole")
greta = createNode(graph, "Person", name="Greta")
kenny = createNode(graph, "Person", name="Kenny")
adam = createNode(graph, "Person", name="Adam")

neo4j = createNode(graph, "Company", name="Neo4j")
digital = createNode(graph, "Company", name="Digital Insights")
docker = createNode(graph, "Company", name="Docker")

createRel(nicole, "WORKS_FOR", neo4j)
createRel(greta, "WORKS_FOR", neo4j)
createRel(kenny, "WORKS_FOR", digital)
createRel(adam, "WORKS_FOR", docker)

createRel(greta, "KNOWS", adam)
createRel(nicole, "KNOWS", kenny)
createRel(kenny, "KNOWS", adam)
```

To create a visualization in visNetwork, you need to pass a nodes data.frame and an edges data.frame. The nodes data.frame requires an `id` column, which is how the node will be identified in the edges data.frame. Optionally, you can pass a `label` column to give the node a text label and a `group` column, where nodes in the same group will be the same color. The edges data.frame requires `from` and `to` columns, with the `id` of the start node and the `id` of the end node for each relationship. These columns should correspond to the `id` column in the nodes data.frame. Optionally, the edges data.frame can also have a `label` column, which will give the edges a text label. Let's write some Cypher queries that will get us all this info in the form we need:

```{r}
node_query = "
MATCH n
RETURN n.name AS id,
       n.name AS label,
       LABELS(n)[0] AS group
"

edge_query = "
MATCH (n)-[r]->(m)
RETURN n.name AS from,
       m.name AS to,
       TYPE(r) AS label
"

nodes = cypher(graph, node_query)
edges = cypher(graph, edge_query)
```

This gives us the following data.frames:

```{r}
nodes
edges
```

Now we can pass these to `visNetwork`:

```{r}
visNetwork(nodes, edges)
```

Nodes can also be sized by a value:

```{r}
neo4j = updateProp(neo4j, employees=1)
digital = updateProp(digital, employees=2)
docker = updateProp(docker, employees=3)

node_query = "
MATCH n
RETURN n.name AS id,
       n.name AS label,
       LABELS(n)[0] AS group,
       n.employees AS value
"

nodes = cypher(graph, node_query)
nodes[is.na(nodes)] = 1

nodes
```

```{r}
visNetwork(nodes, edges)
```

For a more thorough walkthrough of all the styling available, check out visNetwork's [example page](http://dataknowledge.github.io/visNetwork/).

If your graph is much larger, it's more likely you'll want to visualize a small subset of your graph. You can do this by writing a more specific Cypher query, or by getting a random graph. Let's import the movie dataset and go through how we would display a random graph.

```{r}
importSample(graph, "movies", input=F)
```

First, let's get 30 random nodes:

```{r}
LIMIT = 30

node_query = "
MATCH n
WITH n, RAND() AS random
ORDER BY random
LIMIT {limit}
RETURN ID(n) AS id, 
       COALESCE(n.name, n.title) AS label,
       LABELS(n)[0] AS group
"

nodes = cypher(graph, node_query, limit=LIMIT)

head(nodes)
nrow(nodes)
```

Now, in our query for edges, we only want nodes that were found in the previous query:

```{r}
edge_query = "
MATCH (n)-[r]->(m)
WHERE ID(n) IN {ids} AND ID(m) IN {ids}
RETURN ID(n) AS from,
       ID(m) AS to,
       TYPE(r) AS label
"

edges = cypher(graph, edge_query, ids=nodes$id)

head(edges)
```

```{r}
visNetwork(nodes, edges)
```

This gives us a lot of disconnected nodes, unfortunately! We can filter those out, though:

```{r}
nodes$connected = nodes$id %in% c(edges$from, edges$to)
nodes = nodes[nodes$connected, ]

visNetwork(nodes, edges)
```